---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---
## Introduction
Considering the forest fires currently ravaging in Australia, we decided to analyze the continent's weather data and draw conclusions from it, relevant to the phenomena, mainly by predicting rain. The Rain In Australia dataset from Kaggle(https://www.kaggle.com/jsphyg/weather-dataset-rattle-package) was used for this research.


```{r}

library(ggplot2)
library(ggcorrplot)
library(caret)
library(pROC)
```


```{r}
#Preparing data
data = read.csv("weatherAUS.csv", header=TRUE)
tail(data)

#Represent "Yes"/"No" values as numerical data
data$RainTodayNum = as.numeric(data$RainToday)-1
data$RainTomorrowNum = as.numeric(data$RainTomorrow)-1

```
### Data visualization
Plot the data density functions conditioned on the ```RainTomorrow``` variable
```{r}
ggplot(data,aes(x=data$Pressure9am,fill=data$RainTomorrow, stat="count")) + geom_density(alpha=0.5)+
xlab(label = "Pressure at 9 am")+
theme_minimal()

ggplot(data,aes(x=data$Pressure3pm,fill=data$RainTomorrow, stat="count")) + geom_density(alpha=0.5)+
xlab(label = "Pressure at 3 pm")+
theme_minimal()

ggplot(data,aes(x=data$Pressure3pm-data$Pressure9am,fill=data$RainTomorrow, stat="count")) + geom_density(alpha=0.5)+
xlab(label = "Change in pressure")+
theme_minimal()

ggplot(data,aes(x=data$MaxTemp,fill=data$RainTomorrow, stat="count")) + geom_density(alpha=0.5)+
xlab(label = "Maximal temperature")+
theme_minimal()

ggplot(data,aes(x=data$MinTemp,fill=data$RainTomorrow, stat="count")) + geom_density(alpha=0.5)+
xlab(label = "Minimal temperature")+
theme_minimal()

ggplot(data,aes(x=data$MaxTemp-data$MinTemp,fill=data$RainTomorrow, stat="count")) + geom_density(alpha=0.5)+
xlab(label = "Temperature range")+
theme_minimal()

ggplot(data,aes(x=data$Temp9am,fill=data$RainTomorrow, stat="count")) + geom_density(alpha=0.5)+
xlab(label = "Temperature at 9 am")+
theme_minimal()


ggplot(data,aes(x=data$Temp3pm,fill=data$RainTomorrow, stat="count")) + geom_density(alpha=0.5)+
xlab(label = "Temperature at 3 pm")+
theme_minimal()


ggplot(data,aes(x=data$Temp3pm-data$Temp9am,fill=data$RainTomorrow, stat="count")) + geom_density(alpha=0.5)+
xlab(label = "Change in temperature")+
theme_minimal()


ggplot(data,aes(x=data$Humidity9am ,fill=data$RainTomorrow, stat="count")) + geom_density(alpha=0.5)+
xlab(label = "Humidity at 9 am")+
theme_minimal()

ggplot(data,aes(x=data$Humidity3pm ,fill=data$RainTomorrow, stat="count")) + geom_density(alpha=0.5)+
xlab(label = "Humidity at 3 pm")+
theme_minimal()

ggplot(data,aes(x=data$Humidity3pm-data$Humidity9am,fill=data$RainTomorrow, stat="count")) + geom_density(alpha=0.5)+
xlab(label = "Change in humidity")+
theme_minimal()

ggplot(data,aes(x=data$WindGustSpeed ,fill=data$RainTomorrow, stat="density")) + geom_density(alpha=0.5)+
xlab(label = "Wind speed")+
theme_minimal()

ggplot(data,aes(x=data$WindSpeed9am,fill=data$RainTomorrow, stat="count")) + geom_density(alpha=0.5)+
xlab(label = "Wind speed at 9 am")+
theme_minimal()

ggplot(data,aes(x=data$WindSpeed3pm,fill=data$RainTomorrow, stat="count")) + geom_density(alpha=0.5)+
xlab(label = "Wind speed at 3 pm")+
theme_minimal()

ggplot(data,aes(x=data$WindSpeed3pm-data$WindSpeed9am,fill=data$RainTomorrow, stat="count")) + geom_density(alpha=0.5)+
xlab(label = "Change in wind speed")+
theme_minimal()

ggplot(data,aes(x=data$Cloud9am ,fill=data$RainTomorrow, stat="density")) + geom_density(alpha=0.5)+
xlab(label = "Clouds at 9 am")+
theme_minimal()

ggplot(data,aes(x=data$Cloud3pm, fill=data$RainTomorrow, stat="density")) + geom_density(alpha=0.5)+
xlab(label = "Clouds at 3 pm")+
theme_minimal()

ggplot(data,aes(x=data$Cloud3pm-data$Cloud9am, fill=data$RainTomorrow, stat="count")) + geom_density(alpha=0.5)+
xlab(label = "Change in clouds")+
theme_minimal()


ggplot(data,aes(x=data$Sunshine, fill=data$RainTomorrow, stat="density")) + geom_density(alpha=0.5)+
xlab(label = "Sunshine")+
theme_minimal()

ggplot(data,aes(x=data$RainTodayNum, fill=data$RainTomorrow, stat="density")) + geom_density(alpha=0.5)+
xlab(label = "Rain today")+
theme_minimal()

```
From the plots above, distributions of ```WindSpeed9am```, ```WindSpeed3pm``` and ```WindSpeedChange``` do not change considerably depending on ```RainTomorrow```, therefore they can be discarded


### Corelation analysis
Changes throughout the day of some parameters seem to depend on ```RainTomorrow```, so we introduce them as new variables in the dataset.
```{r}
data$TempDiff = data$Temp3pm-data$Temp9am
data$TempRange = data$MaxTemp-data$MinTemp
data$PressureDiff = data$Pressure3pm-data$Pressure9am
data$HumidityDiff = data$Humidity3pm-data$Humidity9am
data$CloudDiff = data$Cloud3pm-data$Cloud9am
```
Plot the correlation matrix for the parameters we are interested in.
```{r}
numerical_data = data[, c("Pressure9am","Pressure3pm","PressureDiff",
                         "MinTemp","MaxTemp","TempRange","Temp9am", "Temp3pm", "TempDiff",
                         "Humidity9am","Humidity3pm","HumidityDiff",
                         "WindGustSpeed",
                         "Cloud9am","Cloud3pm",
                         "RainTodayNum","RainTomorrowNum",
                         "Sunshine")]
numerical_data = na.omit(numerical_data)

ggcorrplot(corr=cor(numerical_data),hc.order = TRUE, method="square", lab = TRUE, lab_size = 2)

```
We choose only values which have absolute value of correlation with ```RainTomorrow``` larger than $0.2$
```{r}

correlated_data = data[, c("Pressure9am","Pressure3pm",
                         "TempRange","TempDiff",
                         "Humidity9am","Humidity3pm",
                         "WindGustSpeed",
                         "Cloud9am","Cloud3pm",
                         "RainTodayNum","RainTomorrowNum",
                         "Sunshine")]
correlated_data = na.omit(correlated_data)

ggcorrplot(corr=cor(correlated_data),hc.order = TRUE, method="square", lab = TRUE, lab_size = 2)



```
### Model preparation

Divide data into training and testing datasets of equal sizes
```{r}
set.seed(29)
training_indices = sample(nrow(correlated_data), nrow(correlated_data)/2, replace=FALSE)
training_data = correlated_data[training_indices, ]
test_data = correlated_data[-training_indices, ]
nrow(training_data)
nrow(test_data)
```

As we try to predict ```RainTomorrow``` which is a categorical, the logistic regression is used 
```{r}
model = glm(training_data$RainTomorrowNum~.,
            family=binomial(link="logit"), data = training_data)
summary(model)
```
From the p-values displayed, ```Pressure9am```, ```Pressure3pm```, ```Humidity3pm```, ```WindGustSpeed```, ```Cloud3pm```, ```RainTodayNum``` and ```Sunshine``` are the most significant ones for rain prediction.

## Testing the model
We use the model to predict ```RainTomorrow``` on the test data, with predicted values that are larger then the threshold of $0.5$ being interpreted as prediction that there will be rain tomorrow. Then, the confusion matrix is ploted to assess the model`s quality. 
```{r}
prediction = predict.glm(model, newdata = test_data, type = 'response')
result = data.frame(prediction = ifelse(prediction > 0.9 , 1 ,0))
result$prediction = factor(result$prediction)
confusionMatrix(result$prediction, factor(test_data$RainTomorrowNum), positive='1')
```
The model was true $85.43\%$ of the time, which is higher than ```No Information Rate```(accuracy of always choosing more frequent value), meaning that the model actually makes valuable predictions.
This is further validated by extremely low P-values and ```Kappa```($\frac{p_{model}-p_{morefreq}}{1-p_{morefreq}}$) of $0.5343$.

The accuracy of negative predictions is $88\%$, meaning that it is pretty safe not to take your umbrella if the model says there will not be rain; it can be increased even more by decreasing the threshold. For example, on the threshold of $0.15$ the confidence level for negative predictions reaches $95.3\%$.

The accuracy of positive predictions is $72.4\%$, meaning that you probably should take your umbrella if the model says there will be rain; it can be increased even more by decreasing the threshold. For example, on the threshold of $0.9$ the confidence level for positive predictions reaches $95.1\%$.